{\small} High Performance P$^{3}$M N-body code: CUBEP3M:

### 2012-05-11 Structure.tex
Upgraded extendedpp.tex, plus minor corrections everywhere.
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
{\bf (Should I mention leapfrog here or anywhere else?)}
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the head processor is active. Other libraries such as {\small P3DFFT} or  Intel {\small MKL-$\alpha$} currently permit this extra level of parallelization,
and it is our plan to migrate to one of these in the near future.

@@

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time {\bf (is this correct?)}. 
The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 
and clears the {\small MPI} communicators. We have collected those operations under the subroutine {\tt finalize} for concision.

>>>>

Each iteration of the main loop starts with the {\tt timestep} subroutine, 
which proceeds to a determination of the redshift jump by comparing the step size constraints from each
force components and from the scale factor.
The cosmic expansion is found by Taylor expanding Friedmann's equation up to the third order in the scale factor,
and can accommodate constant or running equation of state of dark energy.
The force of gravity is then solved  in the {\tt particle\_mesh} subroutine,
which first updates the positions and velocities of the dark matter particles, exchange with neighbouring nodes those that have exited to volume,
creates a new linked list, then solve Poisson's equation.  This subroutine is conceptually identical to that of {\small PMFAST}, 
with the exception  that {\small CUBEP3M} decomposes the volume into cubes (as opposed to slabs). 
The loop over tiles and the particle exchange are thus performed in three dimensions.
{\bf (Should I mention leapfrog here or anywhere else?)}
When the short range and pp forces have been calculated on all tiles, the code exits the parallel {\small OPENMP} loop
and proceeds to the long range. This section of the code is also parallelized in many occasions, but, unfortunately, the current {\small MPI-FFTW}
do not allow multi-threading. There is thus an inevitable loss of efficiency during each global Fourier transforms, during which
only the head processor is active. Other libraries such as {\small P3DFFT} \citep{p3dfft} or  Intel {\small MKL-$\alpha$} \citep{fft_mkl} currently permit this extra level of parallelization,
and it is our plan to migrate to one of these in the near future.

@@

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. Similarly, the code can compute two-dimensional projections of the density field, halo catalogues (see section \ref{sec:halo} for details), and can compute the power spectrum on the coarse grid at run time. 
The code exits the loop when it has reached the final redshift, it then wraps up the {\small FFTW} plans 
and clears the {\small MPI} communicators. We have collected those operations under the subroutine {\tt finalize} for concision.

### 2012-05-11 Systematics.tex
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

One of the most accurate way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
We observe that the agreement with the non-linear prediction of \cite{Lewis} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=3.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The simulation ran on 4000 cores on RANGER.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

@@

The biggest problem with the pp force calculation is that it 
is anisotropic and depends on the location of the fine mesh with respect 
to the particles. An example are two particles on either side of a grid 
cell boundary. These particles will experience the 1-grid cell separation
NGP mesh force, but if the mesh were shifted such that they were
within the same cell, they would experience the (larger) pp force instead. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it can be applied during the particle position update
and suppresses the mesh behavior over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population.
This is convenient if one needs to correlate the initial and final positions of the particles.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the NGP force kernel such as to provide  force as unbalanced as possible at grid cell distances.
This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.

>>>>

One of the most reliable way to assess the simulation's accuracy at evolving particles
is to measure the density power spectrum at late time, and compare to non-linear prediction. 
For an over-density field $\delta({\bf \it x})$, the power spectrum is extracted from the two point function in Fourier space as:
\begin{eqnarray}
\langle | \delta ({\bf \it k}) \delta ({\bf \it k'}) | \rangle = (2\pi)^{3}P(k)\delta_{D}({\bf \it k'} - {\bf \it k})
\label{eq:power}
\end{eqnarray}
where the angle bracket corresponds to an ensemble (or volume) average.
We plot in Fig. \ref{fig:power_highres} the dimensionless power spectrum for a large run ($4000^3$ particles),
which are evolved from a Gaussian initial distribution at $z=100$ down to $z=0$, in a box $3.2 h^{-1}\mbox{Gpc}$ per side.
We observe that the agreement with the non-linear prediction of \cite{Lewis} is at the per cent level over a large dynamical range.
The drop of power at high-$k$ is caused by the finite resolution, and the fluctuations at low-$k$ is caused by the white noise imposed
in the initial conditions.

\begin{figure*}%[ht]
  \begin{center}
    \includegraphics[width=5.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $3.2 h^{-1}\mbox{Gpc}$ per side,
  from $4000^3$ particles. The simulation ran on 4000 cores on Ranger.
    \label{fig:power_highres}}
\end{center}
\end{figure*}

@@

The biggest problem with the pp force calculation is that it 
is anisotropic and depends on the location of the fine mesh with respect 
to the particles. An example are two particles on either side of a grid 
cell boundary. These particles will experience the 1-grid cell separation
NGP mesh force, but if the mesh were shifted such that they were
within the same cell, they would experience the (larger) pp force instead. 
This effect is especially pronounced at the early stages of the simulation where
the density is more homogeneous, and leads to mesh artefacts appearing
in the density field. In order to minimize this systematic effect, 
we randomly shift the particle distribution relative to the mesh by a small
amount -- up to 2 fine grid cells in magnitude -- in each
dimension and at each time step.  This adds negligible computational
overhead as it can be applied during the particle position update
and suppresses the mesh behaviour over multiple time steps.
It is possible to shift back the particles at the end of each time steps,
which prevents a random drift of the whole population.
This is convenient if one needs to correlate the initial and final positions of the particles.

We ensure that, on average, this solution balances out the mesh feature,
by tuning the NGP force kernel such as to provide  force as unbalanced as possible at grid cell distances.
This adjustment is performed from the pairwise force test described in section \ref{sec:accuracy}.
We note that this is one of the driving argument to extend the pp force outside the fine mesh cell,
since the scattering of the NGP force about the actual $1/r^{2}$ law drops rapidly as the distance increases.
As discussed in section \ref{subsec:extendedpp}, this gain in accuracy comes at a price,
a choice that  must be carefully balanced.

At early stages of the simulation, the density fields is rather homogenous, causing the force of gravity to be
rather weak everywhere. In that case, the size of the redshift jumps is controlled by a limit in the cosmological expansion.
If the expansion jump is too large, the size of the residual errors can become significant, and one can observe, for instance,
a growth of structure that does not match the predictions of  linear theory even at the largest scales.
One therefore needs to choose a maximum step size. This is controlled by $r_{max}$, which is the fractional step size,
$\mbox{d}a/(a + \mbox{d}a)$ and is set to 0.05 by default.  It is possible to reduce this number and improve significantly 
the accuracy of the code, at the cost of increasing the total number of time steps. Fig. \ref{fig:ra_max}
shows how $r_{max}$ affects the code. It is a comparison of late time power spectra of density fields that originate from the same initial conditions, 
and used the same random seeds to control the fine mesh shift (mentioned above). We observe that XXXXX.
The {\small CPU} resources required to run these simulations increase rapidly as $r_{max}$ decreases, as seen in Table \ref{table:ra_max}.

a good compromise seems to be XXXXX.

\begin{figure}%[ht]
  \begin{center}
    %\includegraphics[width=3.2in]{graphs/power_highres.eps}
  \caption{Dark matter power spectrum, measured at $z=0$ in a volume $XXX h^{-1}\mbox{Mpc}$ per side,
  from $256^3$ particles and a starting redshift of $z=200$. The different curves show different values of $r_{max}$. 
  The resources required to run these simulations increase rapidly as $r_{max}$ decreases.    \label{fig:ra_max}}
\end{center}
\end{figure}

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the value of $r_{max}$. The tests were performed 
on the CITA Sunnyvale cluster, and general trends could vary slightly on other machines.}
\begin{tabular}{|l|c|c|}
\hline 
$r_{max}$         & time (h)   \\                 
\hline
 $0.1$ & 8.74 \\
 $0.05$ & 11.47\\
 $0.01$ & 14.30 \\
 $0.005$ & 18.87\\
 $0.002$ & 22.52\\
 $0.001$ & 29.62\\
\hline
\end{tabular}
\label{table:ra_max}
\end{center}
\end{table}

@@

### 2012-05-11 paper.tex
Included Ilian's and Vincent's first contribution (Halofinder and f_nl respectively).
Updated references, Intro, scaling, halos with Ilian's input.
Latest changes, including Ilian's latest modifications on halo, non-Gaussian IC, and a few new graphs.

\title[{\small} CUBEP3M: High Performance P$^{3}$M N-body code]{{\small} CUBEP3M: High Performance P$^{3}$M N-body code}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Hugh Merz, Ilian T. Iliev]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Hugh Merz$^{3}$ \thanks{E-mail: merz@sharcnet.ca} and  Ilian T. Iliev$^{4}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Department of XXX\\
$^{4}$Department of XXX\\
$^{5}$Department of XXX}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
\input{Extensions}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3}{}
%\bibliographystyle{amsplain}

>>>>

\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP3M]{{\small} High Performance P$^{3}$M N-body code: CUBEP3M: }
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$ \thanks{E-mail: pen@cita.utoronto.ca}, 
Ilian T. Iliev$^{3}$ \thanks{E-mail: i.t.iliev@sussex.ac.uk} and Hugh Merz$^{4}$ \thanks{E-mail: merz@sharcnet.ca}\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$Department of XXX}

@@

\input{Introduction}
\input{Structure}
\input{PoissonSolver}
\input{ScalingPerformance}
\input{Systematics}
\input{Halos}
\input{Extensions}
\input{Conclusion}
%\input{Appendix}
\bibliographystyle{mn2e}
\bibliography{mybib3_new}{}
%\bibliographystyle{amsplain}
